Geospatial - Assignment 1

### 2025-01-28 hw1/submission_notebook.Rmd
Final commit for paper 4+5 and updated main notebook

# Spatial packages
library(sf)
library(terra) # Newer replacement for raster
library(osmdata) # OpenStreetMap data
library(spData) # Spatial data
library(geobr) # Brazilian boundaries
library(censobr) # Brazilian population data
library(lwgeom) # Needed for st_make_valid

# Data import
library(geodata) # For elevation data
library(readxl) # For Excel files
```

@@

# Paper 4

@@

@@

@@

@@

![Figure x]()

@@

# Paper 5

@@

@@

@@

In comparison the fifth paper map looked as follows:

![Figure x]()

>>>>

# Spatial packages
library(sf)
library(terra) # Newer replacement for raster
library(osmdata) # OpenStreetMap data
library(spData) # Spatial data
library(geobr) # Brazilian boundaries
library(censobr) # Brazilian population data
library(lwgeom) # Needed for st_make_valid
library(rmapshaper) # Simplify and edit spatial features
library(ggrepel) # Repel overlapping text labels in ggplot

# Data import
library(geodata) # For elevation data
library(readxl) # Import and read Excel files
```

@@

# [Paper 4](https://economics.mit.edu/sites/default/files/publications/Catastrophe_Risk_and_Settlement_Location.pdf)

We are looking for a shapefile or dataset specifically containing road network data for Vietnam. We inspected data from GADM (https://gadm.org/data.html#google_vignette) but none of the layers in the file appear to contain road network data. All layers are polygons representing administrative boundaries.

@@

```{r}
# Define the base path
path <- "data/paper4"

# We inspect data from Humdata:
st_layers(file.path(path, "ROADS/vnm_rdsl_2015_OSM.shp"))
# Read the shapefile into an sf object
vietnam_roads <- st_read(file.path(path, "ROADS/vnm_rdsl_2015_OSM.shp"))
view(vietnam_roads)

# Get all the unique values from the column 'type'
unique_values <- unique(vietnam_roads$type)
print(unique_values)
```

Although the data is from 2015, whereas the paper displays roads from 2010, we decide to use this data as road networks.

```{r}
# We inspect data from SimpleMaps:
st_layers(file.path(path,"VIETNAM/"))

vietnam_map <- st_read(file.path(path,"VIETNAM", layer = "vn.shp"))
```

We decide to move forward with this data as Vietnam boundary.

@@

We use the author's documentation (https://www.openicpsr.org/openicpsr/project/207641/version/V1/view;jsessionid=3EBB67FC7920B2090C0D74C349E99F93?path=/openicpsr/207641/fcr:versions/V1/Public-replication-files/Replication_README.pdf&type=file) to understand how he categorised the types of roads.

```{r}
# Categorise roads into group by trying to replicate the author's logic
vietnam_roads <- vietnam_roads %>%
  mutate(
    road_category = case_when(
      type %in% c("motorway", "motorway_link") ~ "Freeway",
      type %in% c("trunk", "trunk_link") ~ "Dual Carriageway",
      type %in% c("secondary") ~ "Major Roads",
      type %in% c("primary", "primary_link", "secondary_link", "tertiary", "tertiary_link", "residential", "living_street") ~ "Minor Roads",
      TRUE ~ "Other Roads"
  )
)
```

@@

```{r}
ggplot() +
  # Adding the map of Vietnam as the base layer, using a light grey fill like the paper
  geom_sf(data = vietnam_map, fill = "lightgrey", color = "black", linetype = "solid", size = 0.5) +

  # Overlaying the road network on the map.
  geom_sf(data = vietnam_roads, aes(color = road_category, linewidth = road_category), alpha = 0.8) +

  # The 'color' aesthetic differentiates road categories
  scale_color_manual(
    values = c(
      "Freeway" = "blue",               
      "Dual Carriageway" = "darkgreen", 
      "Major Roads" = "red",            
      "Minor Roads" = "orange",       
      "Other Roads" = "yellow"        
    )
  ) +
  # The 'linewidth' aesthetic adjusts line thickness based on road types for visual distinction
  scale_linewidth_manual(
    values = c(
      "Freeway" = 1.1,                # Thick line
      "Dual Carriageway" = 1.1,       # Thick line
      "Major Roads" = 0.5,            # Thin line
      "Minor Roads" = 0.5,            # Thin line
      "Other Roads" = 0.3             # Extra thin line
    )
  ) +
  labs(
    title = "Road maps of Vietnam",
    color = "Road Category",
    linewidth = "Road Category"    # Ensures the size legend matches the color legend
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),  # Removes grid lines
    axis.title = element_blank(),  # Removes axis titles
    axis.text = element_blank(),   # Removes axis text
    axis.ticks = element_blank()   # Removes axis ticks
  )
```

@@

![Figure 3](data/paper4/original_plot.png)

@@

# [Paper 5](https://web.stanford.edu/~memorten/ewExternalFiles/Morten_Oliveira_Brasilia.pdf)

The author's replication package is accessible, so we move forward with it's code and data (https://www.openicpsr.org/openicpsr/project/183316/version/V1/view).

@@

```{r}
path <- ("data/paper5/")

# Load the shapefile containing state boundaries for Brazil (1940 projection)
states <- st_read(file.path(path, "GIS_data/uf1940/uf1940_prj.shp"))
# Simplify the state boundaries to reduce file size while maintaining shape integrity
states_simple  <- rmapshaper::ms_simplify(states,keep=0.01,keep_shapes=TRUE)
```

@@

```{r}
# Define the year for the highways data
year  <- 2000

# Construct the file path for the highways shapefile dynamically based on the year
file_name = paste0("/",year,"/highways_",year,"_prj.shp")

# Load the highways shapefile for the specified year
all_highways <- st_read(file.path(path, "GIS_data/roads", file_name))
# Simplify the highways data
all_highways_simple <- rmapshaper::ms_simplify(all_highways,keep=0.01,keep_shapes=TRUE)

# Load and simplify the minimum spanning tree (MST) shapefile for pie distribution
mst_pie <- st_read(file.path(path, "GIS_data/mst/mst_pie_prj.shp"))
mst_pie_simple  <- rmapshaper::ms_simplify(mst_pie,keep=0.01,keep_shapes=TRUE)

# Load and simplify MST data excluding pies for Rio region
mst_all_rio <- st_read(file.path(path, "GIS_data/mst/mst_all_rio_prj.shp"))
mst_all_rio_simple  <- rmapshaper::ms_simplify(mst_all_rio,keep=0.01,keep_shapes=TRUE)

# Load shapefile for capital cities and extract coordinates for labeling
capital_cities <- st_read(file.path(path, "GIS_data/cities/brazil_capital_cities_prj.shp"))
cities_xy <- cbind(capital_cities,st_coordinates(capital_cities))
```

@@

```{r}
# Create a plot showing MST and highway networks in Brazil
fig_5  <- ggplot() +
  # Add state boundaries as the base layer with light grey outlines
  geom_sf(data=states_simple, fill="white", color="grey90") +
  # Add the simplified MST data with a unique linetype and color
  geom_sf(data=mst_pie_simple, size=.6, linetype = "11", aes(color = "Minimum spanning tree"), show.legend = "line")   +
  # Add non-radial highways with a dashed line style
  geom_sf(data=all_highways_simple %>% filter(dm_anlys_p==1 & dm_radial==0), size=.3, linetype = "dashed", aes(color = "Non-radial highways (2000)"), show.legend = "line") +   
  # Add radial highways with a solid line style
  geom_sf(data=all_highways_simple %>% filter(dm_anlys_p==1 & dm_radial==1), size=0.6, aes(color = "Radial highways (2000)"), show.legend = "line") +
  # Minimal theme for simplicity
  theme_minimal() +
  # Overlay capital cities and label them with their names
  geom_point(data=cities_xy, aes(x=X, y=Y)) +
  geom_text_repel(data=cities_xy, aes(x=X, y=Y, label=CITY_NAME)) +
  # Add a custom legend for line types and colors
  labs(color = " ") +
  scale_color_manual(values=c("#777676","#868686","#565555"),
                     guide = guide_legend(override.aes = list(linetype = c("11", "dashed", "solid")))) +
  # Remove axis lines, ticks, and text for a clean map appearance
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )

print(fig_5)
```
In comparison the fifth paper map looked as follows:

![Figure 1](data/paper5/original_plot.png)
